Supporting Information : The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2018-05-19 Paper.Rmd
BMR data calculation completed and description added

As described in Methods, each of our statistics, as computed for the 
ranges $0-0.1$, $0.1-0.2$, ..., $0.9-1$, was compared to the distributions 
of values of the same statistic with the same ranges computed on 1000 normal-copula 
surrogate datasets with similar Kendall or Spearman correlations, in different 
runs for Kendall and Spearman (Fig. \ref{SM-fig_soilCN_nonparam}). Because the correlation 
used in the $\cor$ statistic is the Spearman correlation, Spearman-preserving surrogates
should be used, so the top-right panel of Fig. \ref{SM-fig_soilCN_nonparam}
is more reliable than the top left. Results indicate that tail dependence of 
data is stronger in both the lower and upper tails than would be expected under a 
null hypothesis of a normal copula, answering the second part of question 1. The dependence between C and N is also 
weaker in the middle of the distributions
than would be expected based on a normal-copula null model.

@@

```{r read_BivMS_BMR_data,echo=F}
BivMS_BMR<-readRDS("./Results/BMR_results/BivMS_BMR_lnmass.RDS")
```

@@

The test of indepedence gave $p=$ `r BivMS_BMR$IndepTestRes` (to within the precision available from `BiCopIndTest` in the `VineCopula` package). We fitted several copulas
and got AIC values and upper- and lower-tail dependency estimates based on the fitted copulas (Table \ref{SM-tab_BMRfit}). The table shows that Frank is much better supported (higher AIC and BIC weights) than the Gaussian copula, answering the first part of question 1 for these data.

The model-averaged (AIC weights used for model averaging) lower-tail and upper-tail dependence statistic was $0$ as expected, because the contributions came from Frank and SBB8 which are both symmetric copula. We also tested that lowest-AIC copula Frank was a good-enough fit, giving $p$-value $`r BivMS_BMR$GofRes_CvM`$ (to within the precision available from `BiCopGofTest` in the `VineCopula` package) in the Cramer-von Mises-based test of the goodness of its fit, and $p$-value $`r BivMS_BMR$GofRes_KS`$ in the Kolmogorov-Smirnov-based test of the goodness of fit. The next section looks at tail dependence in a non-parametric way.

#### Non-parametric analysis of tail dependence \label{Ex_BMR_NPA}
<!--we have to add description here after DAN fixes bug(?) in bivfunctionplot.R-->

>>>>

As described in \nameref{Methods}, each of our statistics, as computed for the 
ranges $0-0.1$, $0.1-0.2$, ..., $0.9-1$ (with $10$ number of bins), was compared to the distributions 
of values of the same statistic with the same ranges computed on 1000 normal-copula 
surrogate datasets with similar Kendall or Spearman correlations, in different 
runs for Kendall and Spearman (Fig. \ref{SM-fig_soilCN_nonparam}). 
<!--
Because the correlation 
used in the $\cor$ statistic is the Spearman correlation, Spearman-preserving surrogates
should be used, so the top-right panel of Fig. \ref{SM-fig_soilCN_nonparam}
is more reliable than the top left.
--> 
Results indicate that tail dependence of 
data is stronger in both the lower and upper tails than would be expected under a 
null hypothesis of a normal copula, answering the second part of question 1. The dependence between C and N is also 
weaker in the middle of the distributions
than would be expected based on a normal-copula null model.

@@

```{r read_BivMS_BMR_data,echo=F}
BivMS_BMR<-readRDS("./Results/BMR_results/BivMS_lnBMR_vs_lnmass.RDS")
```

@@

The test of indepedence gave $p=$ `r BivMS_BMR$IndepTestRes` (to within the precision available from `BiCopIndTest` in the `VineCopula` package). We fitted several copulas
and got AIC values and upper- and lower-tail dependency estimates based on the fitted copulas (Table \ref{SM-tab_BMRfit}). The table shows that BB1 and SBB1 are much better supported (higher AIC and BIC weights) than the Gaussian copula, answering the first part of question 1 for these data.

The model-averaged (AIC weights used for model averaging) lower-tail and upper-tail dependence statistic was `r BivMS_BMR$relLTdep_AICw - BivMS_BMR$relUTdep_AICw`, indicating nearly symmetric tail dependence. BB1, being as a two parameter copula,  has both lower and upper tail dependence and depending upon which parameter is greater it shows comparatively greater lower or upper tail dependence. SBB1 is just opposite of BB1 in tail dependency. So, the result was expected, because the contributions came from more or less equally weighted BB1 and SBB1. We also tested that lowest-AIC copula BB1 was a good-enough fit, giving $p$-value `r BivMS_BMR$GofRes_CvM` (to within the precision available from `BiCopGofTest` in the `VineCopula` package) in the Cramer-von Mises-based test of the goodness of its fit, and $p$-value `r BivMS_BMR$GofRes_KS` in the Kolmogorov-Smirnov-based test of the goodness of fit. The next section looks at tail dependence in a non-parametric way.

#### Non-parametric analysis of tail dependence \label{Ex_BMR_NPA}

As described in \nameref{Methods}, each of our statistics, as computed for the ranges $0-0.1$, $0.1-0.2$, ..., $0.9-1$ (with $10$ number of bins), was compared to the distributions of values of the same statistic with the same ranges computed on 1000 normal-copula surrogate datasets with similar Kendall or Spearman correlations, in different runs for Kendall and Spearman (Fig. \ref{SM-fig_BMR_nonparam}).

Results indicate that dependence of data is different in some bins than would be expected under a 
null hypothesis of a normal copula, answering the second part of question 1. But Table \ref{SM-tab_BMR_asym} indicates that tail dependence for the BMR data copula is significantly (95% CI) not different from symmetric normal copula and thus consistent with model selection results.

### 2018-05-19 SupportingInformation.Rmd
BMR data calculation completed and description added

**bmr-contrasts.csv** : This contains the trait contrasts
from phylogenetically independent contrasts [@SukumaranH2010] 
of data reported by [@GenoudIM2017] on the basal metabolic rate (BMR)
and mass of mammals species.
Specifically, the "SELECT" BMR data in their database were used for
 those species for which data were available.
A phylogeny in NEXUS [@NEXUS] format was extracted from the 
 supplementary information of [@GenoudIM2017].
Version 4.4.0 of the DendroPy library [@SukumaranH2010] was used to calculate 
  phylogenetic independent contrasts using Python 2.7.
For each contrast, the path length spanned by the contrast (sum of the branch
  lengths) and the contrast for the mass, natural logarithm of the BMR and natural
  logarithm of the mass were written as a row in the bmr-contrasts.csv file.

@@

# Generating raw data plot btw ln(mass) and BMR from bmr-contrasts.csv
pdf("./Results/BMR_results/rawdataplot_BMR_vs_lnmass.pdf")
op<-par(mar=c(5.1, 4.1, 0.2, 2.1))
plot(BMR_dat$ln.mass,BMR_dat$BMR,type="p",col="red",xlab="ln(mass)",ylab="BMR",
     cex.lab=1.5,cex.axis=1.5)
par(op)
dev.off()
#--------------------------------------------------------------------------------------------
# Generating raw data plot btw ln(mass) and BMR from bmr-contrasts.csv
pdf("./Results/BMR_results/copula_BMR_vs_lnmass.pdf")
v_BMR<-phylo_dat_fn(data=BMR_dat,i=4,j=3,ploton = T) 
#plot(v_BMR[,1],v_BMR[,2],type="p",col="red",xlab="ln(mass)",ylab="BMR",
#     cex.lab=1.5,cex.axis=1.5)
dev.off()
```

\begin{figure}[!h]
\textbf{ \hspace{5.5 cm} (A) \hspace{7 cm} (B)  } \\
\begin{center}
\includegraphics[width=3 in]{./Results/BMR_results/rawdataplot_BMR_vs_lnmass.pdf}
\includegraphics[width=3 in]{./Results/BMR_results/copula_BMR_vs_lnmass.pdf}
\caption[\textbf{(A)} Plot of the raw BMR vs. ln(mass) data \textbf{(B)} Copula plot for BMR dataset]{\textbf{(A)} Plot of the raw BMR vs. ln(mass) data \textbf{(B)} Copula plot for BMR dataset. Obtained by applying ranks separately to each variable, and then dividing the ranks by $n+1$ where $n=`r nrow(v_BMR)`$ is the number of contrasts (see \nameref{MT-Model_selection} of manuscript).\label{fig_raw_cop_BMR}}
\end{center}
\end{figure}

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm]{./Results/stat_results/stat_soilCN/soilCN_bivfunctionplot.pdf}
\caption[Testing non-parametric statistics for soilCN data]{Nonparametric tests for tail dependence and other deviations from a normal copula for \textbf{soil C and N data}. As described in \nameref{MT-Methods}, each statistic was computed on the data using the ranges $0-0.1$, $0.1-0.2$, ..., $0.9-1$ (red lines), and was also computed using the same ranges for 1000 normal-copula surrogate datasets (dashed lines show 0.025 and 0.975 quantiles of these results and dotted lines show 0.005 and 0.995 quantiles).  When red lines were outside dashed lines, text at the top of plots indicates how many surrogate values the data value was less than or greater than. When $\cor$ values (respectively, $\Ps$ or $\Dsq$ values) were greater than surrogates, it means dependence in that part of the distributions was stronger than (respectively, stronger than or weaker than for $\Ps$ or $\Dsq$) expected from a normal-copula null hypothesis. \label{fig_soilCN_nonparam}}
\end{center}
\end{figure}

@@

```{r tab_soilCN_asym,echo=F,results="markup"}
#knitr::opts_knit$set(kable.force.latex = TRUE)
tab_soilCN_asym<-make_table_stat_fn(data=stat_soilCN_ga)
knitr::kable(tab_soilCN_asym, booktabs = T, 
             format = "pandoc",
      #caption.short = "Results for tests of whether asymmetry of tail dependence was significant compared to a normal-copula null hypothesis.",
      caption = "**Soil CN data set results** for tests of whether asymmetry of tail dependence was significant compared to a normal-copula null hypothesis. Three statistics are listed in the left column that measured asymmetry of tail dependence by comparing the distribution ranges $0-0.1$ and $0.9-1$ (Methods). One thousand Kendall-preserving and 1000 Spearman-preserving surrogates with normal copula structure were created, and the same statistics were computed for the surrogates. A table entry $<N$ indicates the value of the given statistic on the data was less than its value on $N$ of the surrogates, so entries of the form $<X$ for $X$ equal to $975$ or above indicate that upper-tail dependence was significantly stronger than lower-tail dependence.\\label{tab_soilCN_asym}")
#kable_styling(, position = "float_left")
```

@@

<!--This chunk is for model selection results of BMR data-->
```{r bivMS_BMR, echo=F, cache=T, cache.extra=list(seed,v_BMR,mtime("OurBiCopSelect.R"),mtime("MyBiCopGofTest.R"),mtime("preprocessors.R"))}
set.seed(seed)
source("OurBiCopSelect.R")
#families<-c(1,3:10,13:14,16:20) 
BivMS_BMR<-OurBiCopSelect(u1=v_BMR[,1],u2=v_BMR[,2],families=c(1,3:10,13:14,16:20),
                          level=0.05,AICBIC="AIC",numBSsmall=100,pthresh=0.2,numBSlarge=1000,gofnormal=F,status=F)
saveRDS(BivMS_BMR,file="./Results/BMR_results/BivMS_BMR_lnmass.RDS")
```

@@

<!--This chunk is for npa testing results of BMR dataset-->
```{r stat_BMR_graphic_approach, echo=F, results="hide", cache=T, cache.extra=list(seed,v_BMR,mtime("bivfunctionplot.R"),mtime("CopulaFunctions_flexible.R"),mtime("copsurrog2d.R"))}
source("bivfunctionplot.R")
set.seed(seed) 
# Now do NPA for v_BMR
# If "F"/"SBB8" is the best fit : then it should be symmetric tail dep (0) : Then still I need to do NPA ?
# May be : because it will also tell me how they are different from "N" copula in different sections?
source("bivfunctionplot.R")
stat_BMR_ga<-bivfunctionplot(v=v_BMR,resloc="./Results/BMR_results/",nametag="BMR_numbin_10",numbin = 10)
stat_BMR_ga_numbin5<-bivfunctionplot(v=v_BMR,resloc="./Results/BMR_results/",nametag="BMR_numbin_5",numbin = 5)
```

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm]{./Results/BMR_results/BMR_numbin_10_bivfunctionplot.pdf}
\caption[Testing non-parametric statistics for BMR data with $numbin=10$]{Nonparametric tests for tail dependence and other deviations from a normal copula for \textbf{BMR data}. As described in \nameref{MT-Methods}, each statistic was computed on the data using the ranges $0-0.1$, $0.1-0.2$, ..., $0.9-1$ (red lines), and was also computed using the same ranges for 1000 normal-copula surrogate datasets (dashed lines show 0.025 and 0.975 quantiles of these results and dotted lines show 0.005 and 0.995 quantiles). \label{fig_BMR_nonparam}}
\end{center}
\end{figure}

>>>>

[**bmr-contrasts.csv**](https://github.com/sghosh89/BIVAN/blob/master/Data/BMR/bmr-contrasts.csv) : 
This contains the trait contrasts
from phylogenetically independent contrasts [@SukumaranH2010] 
of data reported by [@GenoudIM2017] on the basal metabolic rate (BMR)
and mass of mammals species.
Specifically, the "SELECT" BMR data in their database were used for
 those species for which data were available.
A phylogeny in NEXUS [@NEXUS] format was extracted from the 
 supplementary information of [@GenoudIM2017].
Version 4.4.0 of the DendroPy library [@SukumaranH2010] was used to calculate 
  phylogenetic independent contrasts using Python 2.7.
For each contrast, the path length spanned by the contrast (sum of the branch
  lengths) and the contrast for the mass, natural logarithm of the BMR and natural
  logarithm of the mass were written as a row in the bmr-contrasts.csv file.

@@

# Generating raw data plot btw ln(mass) and BMR from bmr-contrasts.csv
pdf("./Results/BMR_results/rawdataplot_lnBMR_vs_lnmass.pdf")
op<-par(mar=c(5.1, 4.1, 0.2, 2.1))
plot(BMR_dat$ln.mass,BMR_dat$ln.BMR,type="p",col="red",xlab="ln(mass) of contrasts",ylab="ln(BMR) of contrasts",
     cex.lab=1.5,cex.axis=1.5)
par(op)
dev.off()
#--------------------------------------------------------------------------------------------
# Generating copula plot btw ln(mass) and ln(BMR) from bmr-contrasts.csv
pdf("./Results/BMR_results/copula_lnBMR_vs_lnmass.pdf")
v_BMR<-phylo_dat_fn(data=BMR_dat,i=4,j=3,ploton = T) 
#plot(v_BMR[,1],v_BMR[,2],type="p",col="red",xlab="ln(mass)",ylab="BMR",
#     cex.lab=1.5,cex.axis=1.5)
dev.off()
```

\begin{figure}[!h]
\textbf{ \hspace{5.5 cm} (A) \hspace{7 cm} (B)  } \\
\begin{center}
\includegraphics[width=3 in]{./Results/BMR_results/rawdataplot_lnBMR_vs_lnmass.pdf}
\includegraphics[width=3 in]{./Results/BMR_results/copula_lnBMR_vs_lnmass.pdf}
\caption[\textbf{(A)} Plot of the raw ln(BMR) of contrasts vs. ln(mass) of contrasts \textbf{(B)} Copula plot for ln(BMR) vs. ln(mass) contrasts]{\textbf{(A)} Plot of the raw ln(BMR) of contrasts vs. ln(mass) of contrasts \textbf{(B)} Copula plot for ln(BMR) vs. ln(mass) contrasts. Obtained by applying ranks separately to each variable, and then dividing the ranks by $n+1$ where $n=`r nrow(v_BMR)`$ is the number of contrasts (see \nameref{MT-Model_selection} of manuscript).\label{fig_raw_cop_BMR}}
\end{center}
\end{figure}

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm]{./Results/stat_results/stat_soilCN/soilCN_bivfunctionplot.pdf}
\caption[Testing non-parametric statistics for soilCN data]{Nonparametric tests for tail dependence and other deviations from a normal copula for \textbf{soil C and N data}. As described in \nameref{MT-Methods}, each statistic was computed on the data using the ranges $0-0.1$, $0.1-0.2$, ..., $0.9-1$ (red cross), and was also computed using the same ranges for 1000 normal-copula surrogate datasets (blue and green cross show 0.025 and 0.975 quantiles of these results).  When red cross was outside blue and green cross, text at the top of plots indicates how many surrogate values the data value was less than or greater than. When $\cor$ values (respectively, $\Ps$ or $\Dsq$ values) were greater than surrogates, it means dependence in that part of the distributions was stronger than (respectively, stronger than or weaker than for $\Ps$ or $\Dsq$) expected from a normal-copula null hypothesis. \label{fig_soilCN_nonparam}}
\end{center}
\end{figure}

@@

```{r tab_soilCN_asym,echo=F,results="markup"}
#knitr::opts_knit$set(kable.force.latex = TRUE)
tab_soilCN_asym<-make_table_stat_fn(data=stat_soilCN_ga)
knitr::kable(tab_soilCN_asym, booktabs = T, 
             format = "pandoc",
      #caption.short = "Results for tests of whether asymmetry of tail dependence was significant compared to a normal-copula null hypothesis.",
      caption = "**Soil CN data set results** for tests of whether asymmetry of tail dependence was significant compared to a normal-copula null hypothesis. Three statistics are listed in the left column that measured asymmetry of tail dependence by comparing the distribution ranges $0-0.1$ and $0.9-1$ (\\nameref{MT-Methods}). One thousand Kendall-preserving and 1000 Spearman-preserving surrogates with normal copula structure were created, and the same statistics were computed for the surrogates. A table entry $<N$ indicates the value of the given statistic on the data was less than its value on $N$ of the surrogates, so entries of the form $<X$ for $X$ equal to $975$ or above indicate that upper-tail dependence was significantly stronger than lower-tail dependence.\\label{tab_soilCN_asym}")
#kable_styling(, position = "float_left")
```

@@

<!--This chunk is for model selection results of BMR data-->
```{r bivMS_BMR, echo=F, cache=T, cache.extra=list(seed,v_BMR,mtime("OurBiCopSelect.R"),mtime("MyBiCopGofTest.R"),mtime("preprocessors.R"))}
set.seed(seed) 
source("OurBiCopSelect.R")
#families<-c(1,3:10,13:14,16:20) 
BivMS_BMR<-OurBiCopSelect(u1=v_BMR[,1],u2=v_BMR[,2],families=c(1,3:10,13:14,16:20),
                          level=0.05,AICBIC="AIC",numBSsmall=100,pthresh=0.2,numBSlarge=1000,gofnormal=F,status=F)
saveRDS(BivMS_BMR,file="./Results/BMR_results/BivMS_lnBMR_vs_lnmass.RDS")
```

@@

<!--This chunk is for npa testing results of BMR dataset-->
```{r stat_BMR_graphic_approach, echo=F, results="hide", cache=T, cache.extra=list(seed,v_BMR,mtime("bivfunctionplot.R"),mtime("CopulaFunctions_flexible.R"),mtime("copsurrog2d.R"))}
source("bivfunctionplot.R")
set.seed(seed) 
# Now do NPA for v_BMR
source("bivfunctionplot.R")
stat_BMR_ga<-bivfunctionplot(v=v_BMR,resloc="./Results/BMR_results/",nametag="BMR_numbin_10",numbin = 10)
stat_BMR_ga_numbin5<-bivfunctionplot(v=v_BMR,resloc="./Results/BMR_results/",nametag="BMR_numbin_5",numbin = 5)
```

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm]{./Results/BMR_results/BMR_numbin_10_bivfunctionplot.pdf}
\caption[Testing non-parametric statistics for BMR data with $numbin=10$]{Nonparametric tests for tail dependence and other deviations from a normal copula for \textbf{BMR data}. As described in \nameref{MT-Methods}, each statistic was computed on the data using the ranges $0-0.1$, $0.1-0.2$, ..., $0.9-1$ (red cross), and was also computed using the same ranges for 1000 normal-copula surrogate datasets (blue and green cross show 0.025 and 0.975 quantiles of these results).  When red cross was outside blue and green cross, text at the top of plots indicates how many surrogate values the data value was less than or greater than. \label{fig_BMR_nonparam}}
\end{center}
\end{figure}
